---
title: "Testing the CAPM"
author: "David Zimmermann"
date: "October 13, 2016"
output: html_document
---

> Warning: due to large datasets (~3Mb) and simulating in total roughly 4,000 portfolios, the compilation of the document may take a while (<2 minute). You can reduce the time by downloading the files and loading the files from your hard drive and/or by reducing the number of simulated portfolios (may introduce in non-max-sharpe portfolios).

# 0. Libraries
```{r, message=FALSE, warning=FALSE}
library(tidyverse) # combines the libraries from the tidyvers
library(magrittr)
library(tseries)
```

# 1.a. Load the Data
```{r, message=FALSE, warning=FALSE}
# Get IRX (risk-free rate) and DJI (Market)
url_dji_selection <- "https://raw.githubusercontent.com/DavZim/Classes/master/Advanced%20Corporate%20Finance/R%20intro/04_capm/data/stock_data.csv"
df_other <- read_csv(url_dji_selection)

# IRX is already reported as returns, therefore rebase to daily returns and calculate
# indexed prices
df_t <- df_other %>% filter(ticker == "IRX") %>% 
  mutate(price = 100 * cumprod(price / (255 * 100) + 1))

df_other <- bind_rows(filter(df_other, ticker == "DJI"), df_t)

# get the stock data for the DJI-stocks
url_dji_all <- "https://raw.githubusercontent.com/DavZim/Classes/master/Advanced%20Corporate%20Finance/R%20intro/04_capm/data/stock_data_large.csv"
df <- read_csv(url_dji_all)
df_all <- bind_rows(df, df_other)

# Calculate the returns (ret)
df_all %<>% group_by(ticker) %>% mutate(ret = price / lag(price) - 1)

# clean the data of missing values to operate with the same data throughout the analysis
df_wide <- df_all %>% select(date, ticker, ret) %>% 
  spread(key = ticker, value = ret) %>% filter(complete.cases(.))

df_all <- df_wide %>% gather(key = ticker, value = ret, -date)

rf <- df_wide %>% summarise(rf = mean(IRX)) %>% as.numeric
exp_ret_m <- df_wide %>% summarise(rf = mean(DJI)) %>% as.numeric

df_all %<>% mutate(date = as.Date(date))

# we only need the variables rf, exp_ret_m, df_all, and df_wide later on, remove the others
rm(df, url_dji_all, url_dji_selection, df_t, df_other)
df_wide
df_all
rf
exp_ret_m
```
# 1.b. Functions
```{r}

#' Calculates the weights in the max-sharpe portfolio
#'
#' The function takes a return matrix and calculates the weights for the max-sharpe portfolio.
#' It does so by simulting n portfolios and then returns the weights of the maximum sharpe portfolio
#' @param ret_mat a matrix with the retuns, one stock per column, each row a date
#' @param shorts a logical value, indicating if short-selling is allowed
#' @param n the number of simulated portfolios, defaults to 1000
#'
#' @return a vector indicating the weights for the max-sharpe portfolio
#'
#' @examples
getWeightsMaxSharpe <- function(ret_mat, shorts, n = 1000) {
  calc_sds <- function(target_returns, stock_mat, shorts = F) {
    ret <- lapply(target_returns, function(x) {
      op <- try(portfolio.optim(stock_mat, x, shorts = shorts)) # to catch errors
      # if unclear about try, see ?try
      
      if (inherits(op, "try-error")) {
        return(NA)
      } else {
        return(op$ps)
      }
    })
    return(unlist(ret))
  }
  
  rets <- seq(from = min(colMeans(ret_mat)), 
              to = max(colMeans(ret_mat)), length.out = n)
  sds <- calc_sds(rets, ret_mat, shorts = shorts)
  
  df <- data_frame(exp_ret = rets, sd_ret = sds) %>% na.omit
  
  df %<>% mutate(sr = (exp_ret - rf) / sd_ret) %>% filter(sr == max(sr))
  
  weights_short <- portfolio.optim(ret_mat, df$exp_ret, shorts = shorts)$pw
  
  return(weights_short)
}
```

# 2. In Sample Test
```{r}
# crate the return matrix
ret_mat <- as.matrix(select(df_wide, -date, -DJI, -IRX))

# find the weights with and without short-selling
weights_short <- getWeightsMaxSharpe(ret_mat, shorts = T, n = 1000)
weights_nt_short <- getWeightsMaxSharpe(ret_mat, shorts = F, n = 1000)

# compute the returns for the portfolios (with and without short-selling)
ret_short <- t(matrix(weights_short, nrow = 1) %*% t(ret_mat))
ret_nt_short <- t(matrix(weights_nt_short, nrow = 1) %*% t(ret_mat))

# combine the data into a data_frame
df_is <- data_frame(date = rep(df_wide$date, 3),
                    type = rep(c("PF_short", "PF_nt_short", "market"), each = nrow(df_wide)),
                    ret = c(ret_short, ret_nt_short, df_wide$DJI))

df_is %<>% group_by(type) %>% mutate(price = 100 * cumprod(1 + ret))
df_is
```

## Plot
```{r}
ggplot(df_is, aes(x = date, y = price, color = type)) + 
  geom_line() + 
  scale_color_manual(name = "Portfolio", labels = c("DJI", "Long-only", "Short"), 
                     values = c("#e41a1c", "#377eb8", "#4daf4a"))
```

# 3. Out of Sample Test
## "Train" the Model (CAPM)
```{r}
# split the dataset into training and testing
df_training <- df_all %>% filter(date < as.Date("2010-01-01"))

# the return matrix for the training dataset
ret_mat_training <- df_training %>% spread(key = ticker, value = ret) %>% 
  select(-date, -DJI, -IRX) %>% as.matrix()

# calculate the weights for each stock
weights_short_oos <- getWeightsMaxSharpe(ret_mat_training, shorts = T, n = 1000)
weights_nt_short_oos <- getWeightsMaxSharpe(ret_mat_training, shorts = F, n = 1000)
```
## "Test" the Model (CAPM) Out Of Sample
```{r}
# take the training dataset
df_test <- df_all %>% filter(date >= as.Date("2010-01-01"))

# the return matrix for the test dataset
ret_mat_test <- df_test %>% spread(key = ticker, value = ret) %>% 
  select(-date, -DJI, -IRX) %>% as.matrix()

# calculate the returns using the weights as estimated before (on the training-set) 
# and the returns from the test dataset
ret_short_oos <- t(matrix(weights_short_oos, nrow = 1) %*% t(ret_mat_test))
ret_nt_short_oos <- t(matrix(weights_nt_short_oos, nrow = 1) %*% t(ret_mat_test))

# include the date and the market (DJI)
df_test_wide <- spread(df_test, key = ticker, value = ret) %>% select(date, DJI)

# combine the data into a single dataset
df_oos <- data_frame(date = rep(df_test_wide$date, 3),
                    type = rep(c("PF_short", "PF_nt_short", "market"), 
                               each = nrow(df_test_wide)),
                    ret = c(ret_short_oos, ret_nt_short_oos, df_test_wide$DJI))

# compute the prices
df_oos %<>% group_by(type) %>% mutate(price = 100 * cumprod(1 + ret))
df_oos
```
## Plot
```{r}
ggplot(df_oos, aes(x = date, y = price, color = type)) + 
  geom_line() + 
  scale_color_manual(name = "Portfolio", labels = c("DJI", "Long-only", "Short"), 
                     values = c("#e41a1c", "#377eb8", "#4daf4a"))
```

# 4. Finding the Mistake

> Question: Does it make sense to have a portolio outperforming the market by roughly 2.5 times?

Answer: Probably not, so where is the mistake in the calculations?

Price: A free beverage at the next bar evening or a snack of your choice from the vending machine in uni during the next class, additionally, eternal fame for doing this! :)