---
title: "R Event Study"
author: "David Zimmermann"
date: '2016-11-23'
output:
  html_document: default
  html_notebook: default
  pdf_document:
    number_sections: yes
    toc: yes
institute: www.github.com/DavZim/
---
\newpage

# Introduction

The following script provides an application of an event study.
The data and and the script itself can be found here: https://github.com/DavZim/Classes/tree/master/Advanced%20Corporate%20Finance/Event%20Study

To run this notebook, make sure that you have at least RStudio version 1.0.44 installed and run R version 3.3.2. 

The general outline for this document is to first load the data, merge the necessary data, inspect the data both visually and via tables, estimate a CMRM (constant-mean-return model), calculate the ARs (abnormal returns), and CARs (cumulative abnormal returns) and then test for significance using a t-test.

# Preparation

## Load libraries
```{r, message=FALSE, warning=FALSE}
library(dplyr)     # for data manipulation
library(ggplot2)   # for plotting
library(lubridate) # for dates
library(readr)     # for data loading
library(scales)    # for plotting
library(tidyr)     # for tidy data
```

## Load data
```{r, message=FALSE, warning=FALSE}
mar_wide <- read_csv("data/market.csv")
ret_wide <- read_csv("data/returns.csv")
events <- read_csv("data/events.csv")

# reshape returns and market to long format
returns <- gather(ret_wide, key = company, 
              value = ret, -date)
market <- gather(mar_wide, key = country, 
                  value = mret, -date)
                  
# date formatting
returns <- returns %>% mutate(date = dmy(date))
market <- market %>% mutate(date = dmy(date))
events <- events %>% mutate(event = dmy(event))
```

## Inspect data
```{r}
returns
market
events
```

## Merge Data 
```{r}
comps <- c("Chrysler", "BellSouth", "Engelhard", "Norsk Hydro", "Pilkington", "INA")
counts <- c("us", "us", "us", "norway", "uk", "italy")
countries <- data_frame(company = comps, country = counts)

# merge into one dataset
merged <- left_join(returns, countries, by = "company")
merged <- left_join(merged, market, by = c("date", "country"))
merged <- left_join(merged, events, by = "company")
merged
```


## Estimation and Events

```{r}

# calculate the event-time as the difference in days to the event
merged <- merged %>% group_by(company) %>% 
  mutate(date_index = 1:n(),
         event_index = max(ifelse(event == date, date_index, 0)),
         event_time = date_index - event_index)

merged
```

Now we want to split our sample into estimation-sample ($[-230, +31]$) and event-sample ($[-30,+30]$).
We also want to have a quick visualization of the return correlations to the market.

```{r}
# windows
estimation_window <- c(-230, -31)
event_window <- c(-30, 30)

# filter returns
estimation <- merged %>% filter(event_time >= estimation_window[1] & 
                                   event_time <= estimation_window[2])

event <- merged %>% filter(event_time >= event_window[1] & 
                              event_time <= event_window[2])

# have a look at the data
estimation
event

# Graph data
theme_set(theme_light())
ggplot(estimation %>% filter(ret != 0), aes(x = mret, y = ret, color = company)) + 
  geom_point() + 
  facet_wrap(~company) +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
   theme(legend.position = "none") +
  labs(title = "Correlations to Market Returns", 
       subtitle = "The respective markets are USA, UK, Norway, and Italy", 
       x = "Market Return", y = "Company Return")
```

# Estimation

## Calculate the CMRM

Although we have many options, this script uses the constant-mean-return model to calculate expected returns (for simplicity reasons mainy). 

The expected return is given by

$$E\left[R_{i,t} \vert X_t\right]$$
using the CMRM (constant mean return model), we get
$$E[R_{i,t} \vert X_t] = \overline{R_{i,t}}$$

```{r}
cmrm <- estimation %>% group_by(company) %>% summarise(cmrm = mean(ret))
cmrm
```

Next, we want to merge the expected returns into the event-dataset to be able to calculate the next steps.

```{r}
# select only necessary variables 
event <- event %>% select(company, ret, event_time)
event <- left_join(event, cmrm, by = "company")
event
```

## Calculate the Abnormal Returns

The abnormal return in period $t$ for company $i$ is given by 
$$AR_{i,t} = R_{i,t} - E\left[R_{i,t}\right]$$
which we can calculate in R like this

```{r}
event <- event %>% mutate(ar = ret - cmrm)
event
```

## Calculate the Cumulative Abnormal Returns

The CARs are given by 
$$ CAR_{i,t}=\displaystyle\sum_{k=1}^t AR_{i,t-k} $$
with a known distribution of
$$ CAR_{i,t} \sim N(0, \sigma_{i,t}^2) $$
In R we can calculate the CARs like this

```{r}
indiv_event <-  event %>% group_by(company) %>% mutate(car = cumsum(ar))
indiv_event

ggplot(indiv_event, aes(x = event_time, y = car, color = company)) + 
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") + 
  geom_step() + 
  scale_y_continuous(labels = percent) +
  labs(title = "Individual Cumulative Abnormal Returns", subtitle = "In the Event-Horizon", 
       x = "Event Time", y = "CAR", color = "Company")
```

We can also calculate aggregated values ($AAR$ as the average abnormal return) per day, which is handy, for example for ploting

```{r}
# aggregated
agg_event <- event %>% group_by(event_time) %>% summarise(aar = mean(ar))
agg_event <- agg_event %>% mutate(car = cumsum(aar))
agg_event

ggplot(agg_event, aes(x = event_time, y = car)) + 
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  geom_step() + 
  scale_y_continuous(labels = percent) +
  labs(title = "Aggregated Cumulative Abnormal Returns", subtitle = "In the Event-Horizon", 
       x = "Event Time", y = "CAR")
```

# Testing

To test for signifance, we mainly use t-test in this script, other tests include Boehmer et al. (1991) and Corrado (1989), among others.

The variance of the $CARs$, are known to be distributed with a variance of
$$\sigma_{i,t}^2 = \frac{1}{N(N-1)} \displaystyle \sum_{j=1}^N \left(CAR_{j,t} - \overline{CAR_{j,t}}\right)^2$$

## T-test

The first chunk uses a t-test to test the individual ARs (the question we are trying to answer: Is the abnormal return in time-period $t$ different from zero?).

```{r}
test1 <- indiv_event %>% 
  group_by(event_time) %>% 
  summarise(mean_ar = mean(ar),
         var_ar = 1/(n()*(n() - 1)) * sum((ar - mean_ar)^2),
         t_value = mean_ar / sqrt(var_ar),
         p_value = pt(abs(t_value), df = n(), lower.tail = F)*2)

test1
```


The following chunk uses CARs to see if the price-development (which is represented by the CARs) is different from zero, instead of a snapshot of a single day as we did in the example above. 

```{r}
# test2 with CARs
stars <- function(p) {
  ifelse(p < 0.001, "***",
         ifelse(p < 0.01, "**",
                ifelse(p < 0.05, "*", "")))
}

test2 <- indiv_event %>% 
  group_by(event_time) %>% 
  summarise(mean_car = mean(car),
         var_car = 1/(n()*(n() - 1)) * sum((car - mean_car)^2),
         t_value = mean_car / sqrt(var_car),
         p_value = pt(abs(t_value), df = n(), lower.tail = F)*2)

test2 %>% mutate(sign = stars(p_value)) %>%
  select(event_time, car = cumsum(mean_car), t_value, sign) %>%
  filter(event_time %in% -3:6) # look only at the frame [-3, 6], to have less output
```

## Testing over Aggregated Times

In the next step we want to look not at a single time-point, but at aggregated times, in this example, we want to see if the price in the time-horizon $[-3, +3]$ is different from zero.

```{r}
time_window <- c(-3, 3)
test3 <- indiv_event %>% filter(event_time >= time_window[1] & 
                                  event_time <= time_window[2]) %>%
  select(company, ar) %>%
  group_by(company) %>% summarise(car = sum(ar)) 

# using the same logic as before
test3 %>% summarise(mean_car = mean(car),
                    var_car = 1/(n()*(n() - 1)) * sum((car - mean_car)^2),
                    t_value = mean_car / sqrt(var_car),
                    p_value = pt(abs(t_value), df = n(), lower.tail = F)*2,
                    sign = stars(p_value))

```

So we can see, that we have detected highly significant returns in the time-period $[-3,+3]$. If we want to test multiple time-periods we can do it like this.
 

## Multiple Time Windows

It may seem a bit more complicated, but we are essentially doing the same thing as before, but use a lapply-function to loop over the row-numbers and repeat the process.

```{r}
time_windows <- data_frame(min = c(-1, 0, -1, -3),
                           max = c(0, 1, 1, 3))

list_events <- lapply(1:nrow(time_windows), function(i) {
  tmp <- indiv_event %>% filter(event_time >= time_windows$min[i] & 
                           event_time <= time_windows$max[i]) %>%
    select(company, ar) %>%
    group_by(company) %>% 
    summarise(car = sum(ar)) %>%
    summarise(mean_car = mean(car),
              var_car = 1/(n()*(n() - 1)) * sum((car - mean_car)^2),
              t_value = mean_car / sqrt(var_car),
              p_value = pt(abs(t_value), df = n(), lower.tail = F)*2,
              sign = stars(p_value)) %>% 
    mutate(range = paste0("[", time_windows$min[i], ", ", 
                          time_windows$max[i], "]"))
  return(tmp %>% select(range, car = mean_car, t_value, p_value, sign))
})
# lapply returns a list of data_frames, to bind them into a single df, we use 
# do.call in combination with rbind.
mult_events <- do.call(rbind, list_events)
mult_events
```

